\documentclass[10pt, a4paper]{article}

% --- Packages ---
\usepackage[top=2cm, bottom=2cm, left=2cm, right=2cm]{geometry}
\usepackage{graphicx} % For images
\usepackage{amsmath}  % For math/equations
\usepackage{hyperref} % For clickable links
\usepackage{caption}  % For styling captions
\usepackage{titlesec} % For controlling section title sizes
\usepackage{float}
\usepackage{xcolor} % Required for defining and using colors
\usepackage{hyperref} % Keep this as your last \usepackage
\usepackage{color}
\usepackage{soul}

% Default highlight color is yellow, you can change it to blue
% \sethlcolor{cyan}

\hypersetup{
    colorlinks=true, % Removes the default boxes and colors the text instead
    linkcolor=blue,  % Color for internal links (e.g., Figure and Table references)
    citecolor=blue,  % Color for bibliographical citations
    urlcolor=blue    % Color for external web URLs
}
% --- Document Formatting ---
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.5em}
\captionsetup{font=scriptsize, labelfont=bf}

% --- Customize Title Sizes ---
\titleformat{\section}
  {\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}
  {\normalsize\bfseries}{\thesubsection}{1em}{}

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{} % Clears default settings
\fancyhead[L]{\footnotesize Weekly Summary \#4 \today} % Top left
\fancyhead[R]{\footnotesize Rachel Broner $\vert$ Miri Adler's Lab} % Top right
\fancyfoot[C]{\thepage} % Bottom middle (page no.)

% \usepackage{mathptmx}

\usepackage{setspace}
\onehalfspacing % For 1.5 line spacing

\begin{document}

\begin{center}
    {\Large \textbf{Weekly Summary \#5 $\vert$ Association Rule Mining}} \\[0.5em]
    \textit{GVHD Analysis} $\vert$ Week of: \today
\end{center}
\vspace{0.2cm}
\hrule
\vspace{0.5cm}

% ---------------------------------------------------------
\section{Current Status \& Highlights}

\subsection*{General Goal}
Our ongoing goal is to mine spatial association rules from MIBI-TOF imaging to map cell-distance dynamics in acute GVHD \cite{ref_gvhd} dataset. 


% ---------------------------------------------------------
\subsection*{This Week's Objectives}
Our primary target this week was:
\begin{enumerate}
    \item \textbf{Validating pipeline's results via Information Theory.} Creating a mathematical framework to "compress" the tissue into information bits and evaluating how accurately our FP-Growth rules describe the tissue's true micro and macro architecture.

    
\end{enumerate}

% ---------------------------------------------------------

\subsection*{Week's Highlights}

\begin{itemize}
\item \textbf{\hl{TBD}} TBD
\end{itemize}

% ---------------------------------------------------------

\section{Next Steps - Primary Focus for the Coming Week:}
Over the next week, we plan to focus on TBD.
\begin{itemize}
    \item \textbf{TBD} TBD.

\end{itemize}

% ---------------------------------------------------------
\vspace{0.5cm}
\hrule
\section{From Tissue to Information: Validating Rules}

\subsection{The Concept: Why Information Theory?}
FP-Growth provides us with isolated local motifs (e.g., 'Cell A is often next to Cell B'). However, the mere existence of a frequent motif does not quantify its actual contribution to the tissue's overall architecture. To measure this 'structural value', we adapt classical Information Theory \cite{Schumacher2015}. Specifically, we measure how much a rule reduces our "surprise" (or error) \cite[Lecture 4, p. 62]{Schumacher2015} when attempting to reconstruct the full tissue map. This information loss is uniquely quantified in bits using \textbf{Kullback-Leibler (KL) Divergence ( $D_{KL}$ )} \cite[Lecture 5, pp. 76-78]{Schumacher2015}.

\subsection{The Problem: Scale and Emergence}
Information theory demonstrates that spatial data can be decomposed into spatial frequencies \cite[Lecture 6, pp. 106-108]{Schumacher2015}:
\begin{itemize}
    \item \textbf{Low Frequencies (Macro):} Highly structured and predictable, these layers possess low entropy and carry minimal "surprise" regarding the tissue's fundamental architecture.
    \item \textbf{High Frequencies (Micro):} These local motifs and precise cell-to-cell interactions carry the high entropy—the true "surprise"—that defines the active biological state of the tissue.
\end{itemize}

If we evaluate our local rules against a naive random baseline, we conflate these scales. The rules "steal credit" by predicting the low-entropy macro-anatomy, artificially inflating their Information Gain. 


\paragraph{The Solution (Multi-Scale Validation):} Analogous to perceptual coding (e.g., JPEG compression), which isolates and prioritizes specific frequencies to measure their true informational value, we must decouple the scales. We establish a baseline ($P_{macro}$) using local spatial permutation \cite{Goltsev2018, Littman2021} that preserves the low-entropy macro-anatomy while intentionally destroying the high-entropy micro-interactions. This isolation enables our \textbf{Emergence Test}, allowing us to evaluate the rules across two scales:
\begin{itemize}
    \item \textbf{Micro-Scale:} Can the rules resolve the high-entropy "surprise" of the local interactions?
    \item \textbf{Macro-Scale:} Can these local micro-rules organically scale up (emerge) to reconstruct the low-entropy global topology, or is the macro-anatomy dictated by top-down external forces?
\end{itemize}

\subsection{The Mathematical Framework: Four Levels of Organization}
To precisely measure information at both scales, we construct four progressively constrained probability distributions, analogous to hierarchical $N^{th}$-order entropy models \cite[Lecture 4, pp. 63-68]{Schumacher2015}. 

\textbf{Critical Safeguard (Laplace Smoothing):} To prevent the mathematical collapse of KL Divergence when encountering unobserved motifs, we apply Laplace Smoothing uniformly across all distributions. We set $\alpha \ll 1$ (e.g., $\alpha = 0.01$) to prevent standard Add-One smoothing from artificially inflating the entropy of rare states \cite[Lecture 4, p. 62]{Schumacher2015}.

\begin{equation}
P(m) = \frac{\text{Count}(m) + \alpha}{N + \alpha \cdot |M|}
\end{equation}
\textit{Where $N$ is the total number of neighborhoods and $|M|$ is the number of possible states in our feature space.}

\paragraph{1. The Absolute Chaos Baseline ($P_{global}$)}
Our "zero-information" baseline. Generated by a completely random global permutation of all cells across the FOV. It destroys both macro-gradients and micro-interactions, while realistically preserving the baseline proportions of the different cell types.

\paragraph{2. The Macro Anatomy Baseline ($P_{macro}$)}
Our Null Model. We shuffle cell locations \textit{locally}. This distribution knows the tissue's "anatomy" (preserving global density gradients) but knows zero interaction rules (destroying micro-motifs). 

\paragraph{3. The Ground Truth ($P_{real}$)}
The actual, empirical distribution of cell motifs in the tissue, generated by counting the exact occurrences of mutually exclusive cell combinations.

\paragraph{4. The Rule-Integrated Model ($P_k$)}
This is where our extracted rules enter the equation. We use \textbf{Iterative Proportional Fitting (IPF)} \cite{Deming1940}. This Maximum Entropy approach effectively predicts global network states based solely on local pairwise interactions \cite{Schneidman2006}. IPF finds the most unbiased probability distribution that perfectly and simultaneously satisfies our set of FP-Growth rule constraints by iteratively updating the states:
\begin{equation}
P_k^{(t+1)}(m) = P_k^{(t)}(m) \times \frac{\text{Rule\_Target\_Probability}}{\text{Current\_Model\_Probability}}
\end{equation}


\subsection{The Ultimate Metric: Multi-Scale Information Gain (IG)}
To prove our rules are biologically meaningful, we measure how much they improve our understanding of the tissue. We do this by calculating the \textbf{Residual Information Gain (IG)}, which is the reduction in error when we upgrade our model from purely anatomical to rule-integrated.

\paragraph{The Core Equations:}
\begin{enumerate}
\item \textbf{Baseline Error ($I_{baseline}$):} How far is the basic anatomical model from reality?
\begin{equation}
I_{baseline} = D_{KL}(P_{real} \parallel P_{macro}) = \sum_{m} P_{real}(m) \log_2 \left( \frac{P_{real}(m)}{P_{macro}(m)} \right)
\end{equation}
\item \textbf{Information Loss ($I_{loss}$):} How far is the new, rule-integrated model from reality?
\begin{equation}
I_{loss} = D_{KL}(P_{real} \parallel P_k) = \sum_{m} P_{real}(m) \log_2 \left( \frac{P_{real}(m)}{P_k(m)} \right)
\end{equation}
\item \textbf{Residual Information Gain (IG):} The actual structural value added by our rules.
\begin{equation}
IG = I_{baseline} - I_{loss}
\end{equation}
\end{enumerate}

\subsection{Multi-Scale Validation (Evaluating Emergence)}

To mathematically test if our local rules accurately explain the micro-environment without corrupting the global tissue macro-architecture, we separate our validation into two distinct Information-Theoretic tests:

\paragraph{Test 1: Macro-Preservation Test}
First, we quantify the total structural information contained strictly in the overarching anatomy relative to absolute chaos:
\begin{equation}
    I_{macro} = D_{KL}(P_{macro} \parallel P_{global}) = \sum_{m} P_{macro}(m) \log_2 \left( \frac{P_{macro}(m)}{P_{global}(m)} \right)
\end{equation}
We then measure the distance between our rule-integrated model ($P_k$) and the global random baseline ($P_{global}$):
\begin{equation}
    I_{macro\_model} = D_{KL}(P_k \parallel P_{global}) = \sum_{m} P_k(m) \log_2 \left( \frac{P_k(m)}{P_{global}(m)} \right)
\end{equation}
\textit{The Test:} If $I_{macro\_model} \approx I_{macro}$, it mathematically proves that enforcing our local micro-rules did not contradict or destroy the natural density gradients of the tissue.

\paragraph{Test 2: Micro-Information Gain Test}
Assuming the macro-structure is preserved, we ask: how much "surprise" did our rules eliminate at the local level? We calculate the Information Loss ($I_{loss}$) between the true tissue and our rule-integrated model:
\begin{equation}
    I_{loss} = D_{KL}(P_{real} \parallel P_k) = \sum_{m} P_{real}(m) \log_2 \left( \frac{P_{real}(m)}{P_k(m)} \right)
\end{equation}
\textit{The Test:} A lower $I_{loss}$ (and thus a higher Information Gain) proves the rules are capturing true biological micro-interactions.

\subsection{Biological Interpretation of the Results}

\begin{itemize}
    \item \textbf{Micro-Victory / Macro-Failure ($I_{loss}$ is low, but $I_{macro\_model}$ diverges significantly from $I_{macro}$):} The rules explain the local neighborhoods perfectly but destroy the global tissue distribution. \textit{Conclusion:} The tissue is governed by top-down external constraints (e.g., blood vessels, strict morphogen gradients) that local cell-to-cell rules do not capture.
    
    \item \textbf{Multi-Scale Victory (Macro is preserved, and $I_{loss}$ is near zero):} The rules reconstruct both the local motifs and the global anatomy. \textit{Conclusion:} The macro-architecture is an emergent property of the local micro-rules. The tissue functions as a self-organizing system.
    
    \item \textbf{Zero Gain ($I_{loss}$ is identical to baseline):} The rules add no new information. \textit{Conclusion:} The extracted rules are merely statistical noise driven by the tissue's existing physical density.
\end{itemize}
% ---------------------------------------------------------
\vspace{0.5cm}
\hrule
\vspace{0.5cm}

% ---------------------------------------------------------


\section{Yet Open Issues (Entering the New Week)}
We enter the next week with yet one main issues in mind:

\begin{itemize}
\item \textbf{Independent Validation:} Find known rules from the literature and see if our pipeline identifies them.

\end{itemize}

% ---------------------------------------------------------


\begin{thebibliography}{9}

\bibitem{ref_gvhd}
Azulay, N., Milo, I., Bussi, Y., Ben-Uri, R., Haran, T. K., Eldar, M., ... \& Keren, L. (2025). A spatial atlas of human gastrointestinal acute GVHD reveals epithelial and immune dynamics underlying disease pathophysiology. \textit{Science Translational Medicine}, 17(823), eadu6032.

\bibitem{Schumacher2015}
Schumacher, B. (2015). \textit{The Science of Information: From Language to Black Holes}. The Great Courses. (Specifically Lectures 4, 5, and 6: Entropy, Surprise, KL Divergence, and Spatial Frequencies).

\bibitem{Goltsev2018}
Goltsev, Y., Samusik, N., Kennedy-Darling, J., Bhate, S., Hale, M., Vazquez, G., ... \& Nolan, G. P. (2018). Deep profiling of mouse splenic architecture with CODEX multiplexed imaging. \textit{Cell}, 174(4), 968-981.

\bibitem{Littman2021}
Littman, R., Hemminger, Z., Foreman, R., Arneson, D., Zhang, G., Gómez-Pinilla, F., Yang, X., \& Wollman, R. (2021). Joint cell segmentation and cell type annotation for spatial transcriptomics. \textit{Molecular Systems Biology}, 17(6), e10108.

\bibitem{Deming1940}
Deming, W. E., \& Stephan, F. F. (1940). On a least squares adjustment of a sampled frequency table when the expected marginal totals are known. \textit{The Annals of Mathematical Statistics}, 11(4), 427-444.

\bibitem{Schneidman2006}
Schneidman, E., Berry, M. J., Segev, R., \& Bialek, W. (2006). Weak pairwise correlations imply strongly correlated network states in a neural population. \textit{Nature}, 440(7087), 1007-1012.

\end{thebibliography}

\end{document}